Commutativity of operation-based CRDTs \\ CIS 673 Project Report

### 2020-04-17 report/draft.tex
Sections 1,2,5.3

\TODO{Motivate CRDTs and why it makes sense to verify them
  automatically.}

@@

\TODO{Describe the technique briefly} \TODO{Describe the tool
  interface. What exactly one has to encode and why}

@@

Say some things about the graph in section 5 of
\cite{shapiro2011conflict} and specification 16 from
\cite{shapiro2011comprehensive}.

\TODO{Describe where we got stuck and why.}

>>>>

Modern day distributed programming presents many challenges. One
key issue that has been well-studied is the idea of maintaining consistency
between replicated instances of data. There are many algorithms and
protocols that present solutions with differing tradeoffs. One major
source of complexity that occurs across various solutions is the idea of
resolving conflicts. If two replicated instances of data receive conflicting
updates separately, how should we specify the merge of these two versions?

Conflict-Free Replicated Data Types (CRDTs) are a recent solution that attempts
to sidestep the issue of merge conflicts. These data structures have semantics
for merging built in to their specification, and ensure that updates made in any
order at any replicas will always commute, and therefore never be in conflict.
However, one downside of these guarantees is the limited sorts of data
structures that have been currently proven to be CRDTs. Furthermore, CRDT
correctness proofs must currently be done manually.

The goal of our project is to reduce the burden of manually verifying that
methods of CRDTs commute (a necessary condition to them being conflict-free). We
employ a tool called Servois that is able to automatically generate
commutativity conditions given pre- and postconditions of an abstract data type.
By helping to automate the verification process of CRDTs, we hope that more data
structures can be proven to be conflict-free, which would allow programmers
richer abstractions they could use to write programs.

@@

Servois \cite{Bansal2018Servois} is a tool that automatically computes the
commutativity (and non-commutativity) conditions of abstract data types (ADTs). Unlike
previous tools that relied on random sampling or blackbox testing, Servois is
able to generate conservative approximations of commutativity conditions that
are guaranteed to be sound, even if they are partial. Servois is
\textit{relatively complete}: if it terminates, the commutativity conditions are
guaranteed to be complete but termination is not guaranteed.

Servois' algorithm can be thought of as similar to counterexample-guided
abstraction refinement. At a high-level, it starts with a coarse partitioning of
the state space and recursively divides it based on increasingly stronger
predicates. Given a partition of the state space corresponding to a certain
predicate, Servois first checks if the ADT is commutative or non-commutative
given that predicate as a precondition on inputs. If it is, no more work is
performed and the predicate is added to the commutativity conditions. If there
are inputs satisfying the predicate that are both commutative and
non-commutative, Servois uses them as counterexamples and partitions the space
recursively. It stops only when the entire state space has been partitioned
completely into commutative or non-commutative sections, or if the user manually
halts the algorithm, in which case Servois returns an overapproximation of the
commutativity conditions.

Servois is an ideal tool to use in this project to alleviate the burden of
performing manual proofs on whether or not methods of a CRDT commute. The only
input it requires is the methods of the datatype whose correctness conditions
are specified as pre- and post-conditions in decidable SMT theories. Since CRDTs
are meant to be specified the same as ordinary ADTs, it is straightforward to
encode them in this format. For our project we have specified the correctness
pre- and postconditions manually, but it would be interesting future work to see
if these invariants could be extracted automatically based on the code.

@@

Our final attempt to encode a CRDT was the directed graph specified in Section 5
of \cite{shapiro2011conflict}. It allows for conflict-free queries, additions,
and deletions of nodes and edges. The specification for the directed graph CRDT
can be seen in \Cref{fig:graph-def}. At a high-level, this graph allows for
simultaneous additions and deletions of vertices and edges across replicas.
These updates are reflected as sets representing the change made by each
operation, and periodically these deltas are sent as updates to all other
replicas. The mechanism by which we are able to handle duplicate
insertions/references is by associating every vertex and edge with a unique tag.

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{directed-graph-def}
    \caption{Op-based Directed Graph Specification}
    \label{fig:graph-def}
\end{figure}

We were able to come up with a straightforward translation of the spec using
CVC4's theory of sets and arrays. However, there were two main issues that
prevented Servois from being able to find commutativity conditions. The
\textit{removeVertex} and \textit{removeArc} (edge) preconditions have an
$\exists$ condition, and CVC4 would immediately return ``unknown'' when queried.
Though we do not know exactly why this occurred, we suspect it is because of the
nested quantification that occurs when introducing an $\exists$ clause in the
specification. Even when we refactored the logic to use a $\forall$, CVC4 still
could not solve the query.

Many of the more complex CRDT specifications contained specifications that
involved quantifiers. It would be interesting to see if these could be
refactored to be quantifier-free. It is important to note that all the example
CRDTs in the Servois paper had quantifier-free encodings. Furthermore, their
method was careful to come up with alternative formulations that avoided
``$\forall\exists$'' quantifier alternation, which they noted would make queries
undecidable \cite{bansal2018servois}.
